---
title: "**Manual basics**"
author: "Jolinde Vlaeyen & Lise Ruitenbeek"
output: html_document
df_print: paged
---

<style type="text/css">
body{
  font-size: 12pt;
  font-family: "Calibri";
}
h1.title{
  font-size: 22pt;
}
h3{
  font-size: 16pt;
}
h4{
  font-size: 12pt;
}
</style>


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


***
### **Index**


***
### **Introduction**
To make this guide accessible for every user, we will also show some very basic R. If you already have experience with R you can probably skip this part.


***
### **Installing R**
R needs to be installed first. We recommend that you also install Rstudio, which is a way to use R with more user interface, so you do not have to code everything yourself but there are some built in buttons to make it easier for you. Below you will find a link to the websites you need to install R and Rstudio, along with a useful guide. Always check if you have the latest version of R.

Guide 
https://rstudio-education.github.io/hopr/starting.html 
Installing R
https://cran.r-project.org/bin/windows/base/ 
Installing Rstudio
https://rstudio.com/products/rstudio/download/ 


***
### **Rstudio interface**
Console, files, ...

Cheat sheet
….


***
### **Updating R**


***
### **Basic R language**
Below we will show you some basics of the R language, and we will refer you to a more elaborate guide. 

Comments # ...
...


***
### **Packages**
R uses so called packages, which you can compare with …. These packages are usually made by other R users and they contain functions that are not a part of the base R. Before you can use a package, you need to install the package. You have to do this only once on your computer, because it will install the package to your computer. But when you use a you can use after installing and loading the package. 


***
### **Help**


***
### **Projects and working directory**



```{r}
```