---
title: "Manual data formatting"
author: "Jolinde Vlaeyen & Lise Ruitenbeek"
output: html_document
df_print: paged
---

<style type="text/css">
body{
  font-size: 12pt;
  font-family: "Calibri";
}
h1.title{
  font-size: 22pt;
}
h3{
  font-size: 16pt;
}
h4{
  font-size: 12pt;
}
</style>


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


***
### **Index**
Different data types (long and wide…)
  How to import data from:
    Excel
    CSV
    Text
    ..

Data types in R

Data structure (variables)

Change data format in R
  Long to wide
  Wide to long


***


https://www.statmethods.net/input/datatypes.html 

The same data can be formatted in different ways. Functions in R, that for example make plots or do statistical tests, expect your data to be in a certain format. It is therefore very important to know how your data is formatted and how it can be changed when necessary. 

Before working in R, you will most likely work with your data in excel. There are however also other 


```{r}
```