---
title: "Manual data formatting"
author: "Jolinde Vlaeyen & Lise Ruitenbeek"
output: html_document
df_print: paged
---

<style type="text/css">
body{
  font-size: 12pt;
  font-family: "Calibri";
}
h1.title{
  font-size: 22pt;
}
h3{
  font-size: 16pt;
}
h4{
  font-size: 12pt;
}
</style>


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


***
### **Index**
Data frames
  Imported
    Excel
    CSV
    Text
    ..
  Created

Data structure 
  Changing data type of variable
  Factors
    Change level order
    Change level names

Useful functions
  Filter and select
  Change data format
    Long to wide
    Wide to long
  Creating tables


***
### **Introduction**
The same data can be formatted in different ways. Functions in R, that for example make plots or do statistical tests, expect your data to be in a certain format. It is therefore very important to know how your data is formatted and how it can be changed when necessary. 


***
### **Data frames**
When working with data in R you will most likely work with so called data frames. Data frames have columns and rows, look a lot like excel files and are easy to work with when doing statistics. When importing your dataset from excel or another format, your data will automatically become a data frame. We will show you how to do this.

**Excel file**

**CSV file**

**Text file**

**Created**


***
### **Data structure and variable type**
After creating a data frame, you need to make sure your columns or variables are in the correct data type. To do this you need to look at the data structure which you can do by clicking on the blue button of the dataset in your right upper corner, or by using the following function:
```{r data structure}
str(dataset)
```

There are five data types or modes that are important to know and your data frame can consist of variables of different types. 
- Numerical ...
- Integer ...
- Character ...
- Factors ...
- Logical ...

You will most likely now intuitively know which type of data these (and other0 variables should be and that they are not correct now. To change the data type of the variables you need to use the following code:
```{r changing data type}
dataset$sex <- factor(dataset$sex)
dataset$age <- integer(dataset$age)
dataset$name <- character(dataset$name)

```

Using as.factor etc....

In case of factors, it is likley that you will once find that the order or the names of the levels make no sense. This can easily be changed.
```{r changing levels}
```


***



```{r}
```