---
title: "**Manual ggplot**"
author: "Jolinde Vlaeyen & Lise Ruitenbeek"
date: "15-02-2021"
output: html_document
df_print: paged
---

<style type="text/css">
body{
  font-size: 12pt;
  font-family: "Calibri";
}
h1.title{
  font-size: 22pt;
}
h3{
  font-size: 16pt;
}
h4{
  font-size: 12pt;
}
</style>


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

***

### **Introduction ggplot ....**

To make plots ... 

The built-in iris dataset will be used ...

Table for which graph to use when ...

***

### **GGplot basics**

With ggplot, there is a specific process to follow. It starts with a basic plot where you tell R which data to use and usually also what variable goes on the x-axis and y-axis for the plot. Then you can add so called "layers" to the plot which for example tell R to make a bargraph or tell R to use another title. 

```{r data, echo=FALSE}
library(ggplot2)
data("iris")

```

```{r ggplot basics, results="hide}
# This is an empty plot, it will not yet know how to plot the data.
ggplot(data = iris)

# You need to add so called aesthetics, with the mapping argument, you can add the variable you want on the x-axis and y-axis, but we will later see that there are much more aesthetics you can add such as colour.
ggplot(data = iris, mapping = aes(x = Sepal.Length, y = Sepal.Width))

# This is the same and will save you some time when making a lot of plots.
ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width))

# The above codes will however still only make an empty graph because R does not now how to represent the data points of the variables. You will need to add a so called geometric object such as geom_point() to make a scatterplot or geom_bar() to make a bargraph. This is called adding a layer, and has to be done by using the +. We will first show a scatterplot example with different possibilities and will later show how to make any other types of plots.
ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width)) + geom_point()



```


