---
title: "**Manual ggplot**"
author: "Jolinde Vlaeyen & Lise Ruitenbeek"
output: html_document
df_print: paged
---

<style type="text/css">
body{
  font-size: 12pt;
  font-family: "Calibri";
}
h1.title{
  font-size: 22pt;
}
h3{
  font-size: 16pt;
}
h4{
  font-size: 12pt;
}
</style>


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


***
### **Index**
Basics
  Data
  Aesthetics
  Layers
  Geometric objects
  
Types of variables to be plotted
  One variable (to look at frequency distribution)
    Numerical 
      Histogram
    Categorical 
      Bar graph 
  Two variables (to look at associations)
    Two numerical 
      Scatter plot
      Line plot (space or time)
    Two categorical 
      Grouped bar plot
      Mosaic plot
    One numerical and one categorical
      Box plot
      Multiple histograms

Extras
  Adding and changing colour
  Shape, size and transparency
  Adding titles
  Changing theme's
  Changing axis 
  Make multiple plots at once


***
### **Introduction ggplot ....**
To make plots ... 
```{r library}
#install.packages("ggplot2")
library(ggplot2)
```
The built-in iris dataset will be used ...

Table for which graph to use when ...

```{r data}
library(readxl)
example_data <- read_excel("example_data.xlsx")

example_data$sex <- factor(example_data$sex)
example_data$individual <- factor(example_data$individual)
```

***
### **GGplot basics**
To create an empty, basic plot: 
```{r empty plot}
ggplot(data = example_data)
```

You need to add so called aesthetics, with the mapping argument, you can add the variable you want on the x-axis and y-axis, but we will later see that there are much more aesthetics you can add such as colour.
```{r plot 1}
ggplot(data = example_data, mapping = aes(x = observation, y = grooming_seconds))
```

This is the same and will save you some time when making a lot of plots.
```{r plot 2}
ggplot(example_data, aes(x = observation, y = grooming_seconds))
```

The above codes will however still only make an empty graph because R does not know how to represent the data points of the variables. You will need to add a so called geometric object such as geom_point() to make a scatterplot or geom_bar() to make a bar graph. This is called adding a layer, and has to be done by using the +. We will first show a scatter plot example with different possibilities and will later show how to make any other types of plots.
```{r plot 3}
ggplot(example_data, aes(x = observation, y = grooming_seconds)) + geom_point()
```

You can also add the aesthetics in the geometric object, and in the case below it will give you the same output. In some cases however it may prove useful to put other aesthetics to the geometric objects than you have already used in the ggplot(), for example when you want to add a prediction line. 
```{r plot 4}
ggplot(example_data) + geom_point(aes(x = observation, y = grooming_seconds))
```


***
###**Types of variables to be plotted**

**Histogram**
```{r histogram}
ggplot(example_data, aes(x = scratching_count)) + geom_histogram(binwidth = 10)
```
You will most likely have to change the binwidth ...


**Bar graph**
```{r bar graph}
ggplot(example_data, aes(x = individual)) + geom_bar()
```
STACKED .....

**Scatter plot**
```{r scatter plot}
```
**Line plot (space or time)**
```{r line plot}
ggplot(example_data, aes(x = observation, y = scratching_count)) + geom_line()
```

**Grouped bar plot**
To make a grouped bar plot, we need to use an extra layer, which is called faceting. DIT IS GEEN LOGISCH VOORBEELD 
```{r grouped bar plot}
ggplot(example_data, aes(x = individual)) + geom_bar() + facet_wrap(~sex)

```
**Mosaic plot**
```{r mosaic plot}

```

**Box plot**
```{r boxplot}

```
**Multiple histograms**
```{r multiple histogram}
```


***
###**Extra's**
**Adding and changing colour**
```{r colour}
ggplot(example_data, aes(x = observation, y = grooming_seconds, color = individual)) + geom_point()
```

When adding colour to a bargraph, you most likely will need to use the "fill" statement in stead of the "color" statement. See the difference:
```{r colour bargraph}
ggplot(example_data, aes(x = individual, color = individual)) + geom_bar()

ggplot(example_data, aes(x = individual, fill = individual)) + geom_bar()

```

R gives your graph a standard colour, but you can easily change this. When you want all your points to have the same (but different than the standard) colour, you can do this:
```{r change color}
ggplot(example_data, aes(x = observation, y = grooming_seconds)) + geom_point(color = "blue")
```
Note that this "color" statement has to be within the geom and without the aes(). You can use any existing colour from by name or by code. See the following link for all colour name codes...

There are futhermore two ways to change the colour of different groups in your graph. The first one is to chose a palette, from the link below ... . You can then add the following to your code:
```{r colour palette}
ggplot(example_data, aes(x = observation, y = grooming_seconds, color = individual)) + geom_point() + scale_color_brewer(palette = "Dark2")
```
When you do this for a bargraph and you used the "fill" statement in stead of the "color" statement, you should use:
```{r fill palette}
ggplot(example_data, aes(x = individual, fill = individual)) + geom_bar() + scale_fill_brewer(palette = "Dark2")
```

The second way to change the colours of your graph is to do this manually. For this you need to look up the colour names or codes (see link....). Instead of using "scale_color_brewer()" or "scale_fill_brewer()", you will now have to use "scale_color_manual()" or "scale_fill_manual()".
```{r color manual}
ggplot(example_data, aes(x = observation, y = grooming_seconds, color = individual)) + geom_point() + scale_color_manual(values = c("#999999", "#E69F00", "#56B4E9"))
```

```{r fill manual}
ggplot(example_data, aes(x = individual, fill = individual)) + geom_bar() + scale_fill_manual(values = c("red", "green", "blue"))
```
As you can see, you can use the colour code or the colour name.


**Shape, size and transparency**
Besides changing the colour, it is also possible to change the shape, size and transparency of your figure to make it more understandable. 

Like colour, you can also manually change shape, size and transparency of all your points in your plot. Maybe you want squares instead of points in your scatterplot. You can check all possible shapes in the link below...
```{r shape}
ggplot(example_data, aes(x = observation, y = grooming_seconds)) + geom_point(shape = 0)
```

You can also for example increase the size of your points when there are only a few. 
```{r size}
ggplot(example_data, aes(x = observation, y = grooming_seconds)) + geom_point(size = 5)
```

Or you can increase the transparency when there are a lot of points clustered together. ANDER VOORBEELD
```{r trans}
ggplot(example_data, aes(x = observation, y = grooming_seconds)) + geom_point(alpha = 0.1)
```

Furthermore, changing the shape, size or transparency of your points in a scatterplot can have a similar effect as colour when it is used for different groups:
```{r shape 2}
ggplot(example_data, aes(x = observation, y = grooming_seconds, shape = individual)) + geom_point()
```
```{r size 2}
ggplot(example_data, aes(x = observation, y = grooming_seconds, size = individual)) + geom_point()
```
```{r trans 2}
ggplot(example_data, aes(x = observation, y = grooming_seconds, alpha = individual)) + geom_point()
```
Size and transparency can also be used for numerical variables which gives it an extra advantage. 

R again gives your groups a standard shape, size and transparency but again this can easily be changed. 
```{r shape 3}
ggplot(example_data, aes(x = observation, y = grooming_seconds, shape = individual)) + geom_point() + scale_shape_manual(values = c(3, 6, 12))
```
```{r size 3}
ggplot(example_data, aes(x = observation, y = grooming_seconds, size = individual)) + geom_point() + scale_size_manual(values = c(2,7,10))
```
```{r trans 3}
ggplot(example_data, aes(x = observation, y = grooming_seconds, alpha = individual)) + geom_point() + scale_alpha_manual(values = c(0.8, 0.5, 0.1))
```


**Adding titles**


**Changing theme's**
**Changing axis** 
**Make multiple plots at once**


